---
title: "Practical Machine Learning Course Project"
author: "Jitendra Patel"
output: html_document
---

### Objective: 

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

### Goal and Submission
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

#### Reproducibility and Libraries
In order to reproduce the same results, you need a certain set of packages, as well as setting a pseudo random seed.
````{r, eval=FALSE}
library(caret)
library(randomForest)
````
#### Getting Data and Loading to memory
````{r}
trainUrl <- "http:///d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

trainingRaw <- read.csv(file="pml-training.csv", header=TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep=',', na.strings=c('NA','','#DIV/0!'))
testingRaw <- read.csv(file="pml-testing.csv", header=TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep=',', na.strings=c('NA','','#DIV/0!'))

trainingRaw$classe <- as.factor(trainingRaw$classe)  
````

#### Cleaning variables
After investigating all the variables of the sets, it's possible to see that there are a lot of values NA or useless or empty variables for the prediction. It's request to compute the prediction only on the accelerometers values of belt, forearm, arm and dumbell. So, the non-accelerometer measures are discard with the useless variables.

````{r, eval=FALSE}
NAindex <- apply(trainingRaw,2,function(x) {sum(is.na(x))}) 
trainingRaw <- trainingRaw[,which(NAindex == 0)]
NAindex <- apply(testingRaw,2,function(x) {sum(is.na(x))}) 
testingRaw <- testingRaw[,which(NAindex == 0)]
````
#### Preprocessing variables
````{r, eval=FALSE}

v <- which(lapply(trainingRaw, class) %in% "numeric")

preObj <-preProcess(trainingRaw[,v],method=c('knnImpute', 'center', 'scale'))
trainLess1 <- predict(preObj, trainingRaw[,v])
trainLess1$classe <- trainingRaw$classe

testLess1 <-predict(preObj,testingRaw[,v])
````
#### Removing the non zero variables
Removing the variables with values near zero, that means that they have not so much meaning in the predictions
````{r, eval=FALSE}
nzv <- nearZeroVar(trainLess1,saveMetrics=TRUE)
trainLess1 <- trainLess1[,nzv$nzv==FALSE]

nzv <- nearZeroVar(testLess1,saveMetrics=TRUE)
testLess1 <- testLess1[,nzv$nzv==FALSE]
````

#### Create cross validation set
The training set is divided in two parts, one for training and the other for cross validation

````{r, eval=FALSE}
set.seed(12356)

inTrain = createDataPartition(trainLess1$classe, p = 3/4, list=FALSE)
training = trainLess1[inTrain,]
crossValidation = trainLess1[-inTrain,]
````

#### Train model
Train model with random forest due to its highly accuracy rate. The model is build on a training set of 28 variables from the initial 160. Cross validation is used as train control method.
````{r, eval=FALSE}
modFit <- train(classe ~., method="rf", data=training, trControl=trainControl(method='cv'), number=5, allowParallel=TRUE )

````
````
Random Forest 

14718 samples
   27 predictor
    5 classes: 'A', 'B', 'C', 'D', 'E' 

No pre-processing
Resampling: Cross-Validated (10 fold) 
Summary of sample sizes: 13246, 13248, 13245, 13247, 13246, 13245, ... 
Resampling results across tuning parameters:

  mtry  Accuracy   Kappa      Accuracy SD  Kappa SD   
   2    0.9931375  0.9913192  0.001517843  0.001919988
  14    0.9928663  0.9909766  0.003092620  0.003911705
  27    0.9902162  0.9876245  0.002972839  0.003758753

Accuracy was used to select the optimal model using  the largest value.
The final value used for the model was mtry = 2. 
````

#### Accuracy on training set and cross validation set
Following the computation on the accuracy of trainig and cross validation set

Training set:
````{r, eval=FALSE}
trainingPred <- predict(modFit, training)
confusionMatrix(trainingPred, training$classe)
````
````
Confusion Matrix and Statistics

          Reference
Prediction    A    B    C    D    E
         A 4185    0    0    0    0
         B    0 2848    0    0    0
         C    0    0 2567    0    0
         D    0    0    0 2412    0
         E    0    0    0    0 2706

Overall Statistics
                                     
               Accuracy : 1          
                 95% CI : (0.9997, 1)
    No Information Rate : 0.2843     
    P-Value [Acc > NIR] : < 2.2e-16  
                                     
                  Kappa : 1          
 Mcnemar's Test P-Value : NA         

Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
Sensitivity            1.0000   1.0000   1.0000   1.0000   1.0000
Specificity            1.0000   1.0000   1.0000   1.0000   1.0000
Pos Pred Value         1.0000   1.0000   1.0000   1.0000   1.0000
Neg Pred Value         1.0000   1.0000   1.0000   1.0000   1.0000
Prevalence             0.2843   0.1935   0.1744   0.1639   0.1839
Detection Rate         0.2843   0.1935   0.1744   0.1639   0.1839
Detection Prevalence   0.2843   0.1935   0.1744   0.1639   0.1839
Balanced Accuracy      1.0000   1.0000   1.0000   1.0000   1.0000
````

Cross validation set
````{r, eval=FALSE}
cvPred <- predict(modFit, crossValidation)
confusionMatrix(cvPred, crossValidation$classe)
````
````
Confusion Matrix and Statistics

          Reference
Prediction    A    B    C    D    E
         A 1395    2    0    0    2
         B    0  946    2    0    0
         C    0    1  850    6    0
         D    0    0    3  797    3
         E    0    0    0    1  896

Overall Statistics
                                          
               Accuracy : 0.9959          
                 95% CI : (0.9937, 0.9975)
    No Information Rate : 0.2845          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.9948          
 Mcnemar's Test P-Value : NA              

Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
Sensitivity            1.0000   0.9968   0.9942   0.9913   0.9945
Specificity            0.9989   0.9995   0.9983   0.9985   0.9998
Pos Pred Value         0.9971   0.9979   0.9918   0.9925   0.9989
Neg Pred Value         1.0000   0.9992   0.9988   0.9983   0.9988
Prevalence             0.2845   0.1935   0.1743   0.1639   0.1837
Detection Rate         0.2845   0.1929   0.1733   0.1625   0.1827
Detection Prevalence   0.2853   0.1933   0.1748   0.1637   0.1829
Balanced Accuracy      0.9994   0.9982   0.9962   0.9949   0.9971
````

## Results and Files to submit as Answers:
Finally, here is the Predictions on the real testing set
````{r, eval=FALSE}
testingPred <- predict(modFit, testLess1)
````
Function to generate files with predictions to submit Answers for Assignment:
````{r, eval=FALSE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(testingPred)
````